Early Warning Signals and the Prosecutor's Fallacy

### 2013-03-04 prosecutor/fallacy.tex
minor

\usepackage{endfloat} % Move floats to the end of the document
\renewcommand{\listoffigures}{} % but supress these lists
\renewcommand{\listoftables}{} % supress these lists
% compile without figures?
\newif\ifhavefigures
\havefigurestrue
\newif\ifnohavefigures % allows alternate condition
\nohavefigurestrue     % turn on the alternate condition uncommenting this

@@

\begin{figure}
  \begin{center}
      \ifhavefigures
%        \includegraphics[width=\textwidth]{Figure1}
      \fi
      \ifnohavefigures
        \hspace{\textwidth}
      \fi
      \label{fig:bifurcation}
  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. 
\textbf{(a)} Plot of the model functions shown in Eq~\eqref{master}
with parameters $a=180$, $K=500$, $e=.5$, and $h=200$.  When the death
rate is higher than the birth rate, the system dynamics drive the state
(population size) to smaller values.  When birth rate is higher, the
system moves right, as indicated by the arrows.\textbf{(b)} The potential
energy is given by the negative integral of $b(n)-d(n)$, shown in the
lower plot.   The potential function gives an intuitive picture of
the stability of a system by imagining the curve as a surface on which
a ball is free to bounce across -- wells correspond to stable points
and peaks to unstable points. While most trajectories remain near the
stable well, some transition out merely by chance.  An example of such
a trajectory is shown in the top panel, in which time increases along
the vertical axis. Though initially oscillating around the stable state,
a chance excursion carries it beyond the threshold (vertical dotted line).
Such chance trajectories can produce the statistical patterns as observed
in true critical transitions seen in panel \textbf{(c)}: Early warning
signals  are aimed at detecting systems which are slowly moving towards
a tipping point or bifurcation, illustrated in the successive curves
(detiorating and critical). Top panel: An example trajectory from a
simulation under this process shows the state of the system as the
potential moves towards the bifurcation point.
} \label{fig:1}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \ifhavefigures
%    \includegraphics[width=6in]{Figure2}
    \fi
     \ifnohavefigures
     \hspace{6in}
    \fi
 \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

\begin{figure}
  \begin{center}
    \ifhavefigures
%    \includegraphics[width=6in]{Figure3}
    \fi
    \ifnohavefigures
     \hspace{6in}
    \fi
  \end{center}
  \caption{The identical analysis from Figure~\ref{fig:indicator} is shown for the model in Eq~\eqref{May1977} using parameters $r = 0.75$, $K=10$, $a=1.7$, $Q=3$, and $H=1$. A similar statistical bias, particularly towards positive values of $tau$ occurs in this model as well.}
  \label{fig:may}
\end{figure}

@@

 \section{Acknowledgements}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department of
Energy grant DE-FG02-97ER25308 to CB. The authors thank M. Baskett, 
T.A. Perkins and N. Ross for helpful comments on earlier drafts of the
manuscript, and also P. Ditlevsen and an anonymous reviewer for their 
comments.

>>>>

\usepackage{endfloat} % Move floats to the end of the document
\renewcommand{\listoffigures}{} % but supress these lists
\renewcommand{\listoftables}{} % supress these lists
% compile without figures?
\newif\ifhavefigures
\havefigurestrue
\newif\ifnohavefigures % allows alternate condition
%\nohavefigurestrue     % turn on the alternate condition uncommenting this

@@

\begin{figure}
  \begin{center}
      \ifhavefigures
        \includegraphics[width=\textwidth]{Figure1}
      \fi
      \ifnohavefigures
        \hspace{\textwidth}
      \fi
      \label{fig:bifurcation}
  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. 
\textbf{(a)} Plot of the model functions shown in Eq~\eqref{master}
with parameters $a=180$, $K=500$, $e=.5$, and $h=200$.  When the death
rate is higher than the birth rate, the system dynamics drive the state
(population size) to smaller values.  When birth rate is higher, the
system moves right, as indicated by the arrows.\textbf{(b)} The potential
energy is given by the negative integral of $b(n)-d(n)$, shown in the
lower plot.   The potential function gives an intuitive picture of
the stability of a system by imagining the curve as a surface on which
a ball is free to bounce across -- wells correspond to stable points
and peaks to unstable points. While most trajectories remain near the
stable well, some transition out merely by chance.  An example of such
a trajectory is shown in the top panel, in which time increases along
the vertical axis. Though initially oscillating around the stable state,
a chance excursion carries it beyond the Allee threshold (vertical dotted line).
Such chance trajectories can produce the statistical patterns as observed
in true critical transitions seen in panel \textbf{(c)}: Early warning
signals  are aimed at detecting systems which are slowly moving towards
a tipping point or bifurcation, illustrated in the successive curves
(deteriorating and critical). Top panel: An example trajectory from a
simulation under this process shows the state of the system as the
potential moves towards the bifurcation point. The original position of the Allee threshold
is shown by the vertical dotted line (though it moves slightly as the parameter changes).  
} \label{fig:1}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \ifhavefigures
    \includegraphics[width=6in]{Figure2}
    \fi
     \ifnohavefigures
     \hspace{6in}
    \fi
 \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

\begin{figure}
  \begin{center}
    \ifhavefigures
    \includegraphics[width=6in]{Figure3}
    \fi
    \ifnohavefigures
     \hspace{6in}
    \fi
  \end{center}
  \caption{The identical analysis from Figure~\ref{fig:indicator} is shown for the model in Eq~\eqref{May1977} using parameters $r = 0.75$, $K=10$, $a=1.7$, $Q=3$, and $H=1$. A similar statistical bias, particularly towards positive values of $tau$ occurs in this model as well.}
  \label{fig:may}
\end{figure}

@@

 \section{Acknowledgments}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department
of Energy grant DE-FG02-97ER25308 and NERSC Supercomputing grant
DE-AC02-05CH11231 to CB. The authors thank M. Baskett, T.A. Perkins
and N. Ross for helpful comments on earlier drafts of the manuscript,
and also P. Ditlevsen and an anonymous reviewer for their comments.
